Climate indicators

### 2023-03-06 DRAFT_climate_indicators.Rmd
clean text

*Author and date:* 
Anders L. Kolstad
March 2023

@@

### Aditional comments about the dataset

The data format has changes from .BIL to .nc (netcdf) and now a single file contains all the rasters for one year (365 days), and sometimes for multiple variables also.

@@

### Norwegain standard

@@

The reference condition for climate variables is defined as the laste normal period 1961-1990.

@@

-   Un-scaled indicator value = median value over last 5 year (moving window)

@@

-   Threshold for good ecosystem condition = 2 standard deviation units for the climate variable during the reference period.

-   Lower reference value = 5 standard deviation units for the climate variable during the reference period.

@@

For the indicator map (1 x 1 km raster) there is no uncertainty associated with the indicator values. For aggregated indicator values (e.g. for regions), the uncertainty in the indicator value is calculated from the spatial variation in the indicator values via bootstrapping.

@@

rr and tm are being download from <https://thredds.met.no/thredds/catalog/senorge/seNorge_2018/Archive/catalog.html>

@@

```{r}
senorgelayers <- read_delim("data/senorgelayers.txt", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
DT::datatable(senorgelayers)
```

@@

I think I know the CRS, so setting it manually.

@@

Checking that this is correct
```{r}
p = st_sfc(st_point(c(7039947.63, 270783.69)), crs=25833)
```

@@

Initially the data had four attributes, but I subsettet to only include _tg_.

@@

Saving these to file to demonstrate the importing step of the workflow.

@@

Here is some dummy data for accounting areas (regions) and ecosystem occurrences as well.

@@

```{r}
nor <- sf::st_read("data/outlineOfNorway_EPSG25833.shp")
```
Remoce marine areas from regions

@@

```{r, fig.cap="Five accounting areas (regions)."}

@@

@@

### Conseptual workflow

The general, conceptual workflow is like this:

@@

    -   Import .nc files (loop though year 1961-1990) and subset to the correct attribute (above)

    -   Filter data by dates (optional) (`dplyr::filter`)

    -   Aggregate across time within a year (`stars::st_apply`)

@@

4.  Write to file

5.  Mask by ecosystem type (this could also be done in step one to save time)

6.  Aggregate in space (to accounting areas)

@@

    -   Intersect with accounting area polygons `sf:st_intersects`

    -   Bootstrap indicator values (cell values) and get median and sd

7.  Make trend figure

I will try to use [parallelization](https://tmieno2.github.io/R-as-GIS-for-Economists/EE.html) to speed things up. Also, masking to ecosystem time earlier in the workflow might prove smart.

@@

@@

For this example I want to calculate the mean temperature over the summer. I therefore need to aggregate over time. There are two ways, either using `st_apply` or using `aggregate`.

@@

The *v\_* follows the naming convention in Norway.

@@

```{r}
ggplot()+
  geom_stars(data = summerTemp_fullSeries) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~Year)
```

@@

_I initially tried to read stars proxy object and add steps to the call_list and execute these with `st_as_stars`, but this still returned proxy objects. Therefore I read the whole file in without proxy. To save time I can subset based on Julian dates, but this is a hard coded solution._

@@

myFiles <- list.files(path2, pattern=".nc$",full.names = T)
# The last file, the last year, is incomplete and dont include all juliani dates that we select below
myFiles <- myFiles[-length(myFiles)]
real_temp_summer <- NULL

cl <- makeCluster(10L)

# Get julian days
temp <- stars::read_ncdf(paste(myFiles[1]), var="tg")
start_month_num <-  6
end_month_num <- 8

@@

  tic("init")
  temp <- stars::read_ncdf(paste(myFiles[i]), var="tg", proxy=F,
                           ncsub = cbind(start = c(1, 1, julian_start), 
                              count = c(NA, NA, step)))
  year_temp <- year(st_get_dimension_values(temp, "time")[1])
  print(year_temp)
  #start <- ym(paste(year_temp, start_month, sep="-"))
  #end <-  ym(paste(year_temp, end_month, sep="-"))
  #myInterval <- interval(start, end)
  #myInterval <- interval(start, start+1)
  lookup <- setNames("mean", paste0("v_", year_temp)) # Perhaps leave out the v_ to get a numeric vector instead, which is easier to subset
  st_crs(temp) <- 25833
  toc()

@@

This takes about 20 sec per file/year, or 22 min on total. Here's a test for the effect of splitting over more cores.
```{r, eval=F}
cl <- makeCluster(2L)
cl2 <- makeCluster(6L)
cl3 <- makeCluster(10L)

@@

```
No cluster: 21.401 sec elapsed
Two clusters: 21.546 sec elapsed
Six clusters: 14.548 sec elapsed
Ten clusters: 15.29 sec elapsed
Ten clusters (second run): 13.617 sec elapsed

@@

These files are about 500 MB, but the tiff file reads in a lot quicker then the rds file, so from now on I'll just export as tiff.

@@

Note that GTiff automatically renames the third dimension _band_ and also renames the attribute.

@@

I can rename them. 
```{r}
temp_median_summer <- temp_median_summer %>% 
  st_set_dimensions(names = c("x", "y", "v_YEAR")) %>%
  setNames("temperature")
dim(temp_median_summer)
```

@@

### 2. Calculate reference values
#### Aggregate across reference years

@@

```{r}
st_get_dimension_values(y, "time")
```
```{r}
y_filtered <- y %>%
  select(Attribute_A) %>%
  filter(time %within% interval("1961-01-01", "1990-12-31"))
```

@@

This is also something I can export.
```{r, eval=F}
# write_stars can only export a single attribute 
#write_stars(temp_median_summer_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.tiff")

@@

system.time(
y_var_norm <- y_var %>%
  mutate(reference_low = reference_upper - 5*sd ) %>%
  mutate(reference_low2 = reference_upper + 5*sd ) %>%
  mutate(threshold_low = reference_upper -2*sd ) %>%
  mutate(threshold_high = reference_upper +2*sd ) %>%
  mutate(across(all_of(cols), ~ 
                  if_else(.x < reference_upper,
                  if_else(.x < threshold_low, 
                                        (.x - reference_low) / (threshold_low - reference_low),
                                        (.x - threshold_low) / (reference_upper - threshold_low),
                                        ),
                  if_else(.x > threshold_high, 
                                        (reference_low2 - .x) / (reference_low2 - threshold_high),
                                        (threshold_high - .x) / (threshold_high - reference_upper),
                                        )
                ))) %>%
  mutate(across(all_of(cols), ~ if_else(.x > 1, 1, .x))) %>%
  mutate(across(all_of(cols), ~ if_else(.x < 0, 0, .x))) %>%
  rename(all_of(cols_new)) %>%
  c(select(y_var, all_of(cols)))
)
```
  user  system elapsed 
 14.803   2.717  17.512

@@

The _real_ indicator values should be running means over 5 years, but calculating a running mean for all time steps is too time consuming. Therefore, the time series will be presented with yearly resolution, and the scaled, regional indicator values will be calculated for distinct time steps.

>>>>

*Author and date:*

Anders L. Kolstad

March 2023

@@

### Additional comments about the data set

The data format has recently changed from .BIL to .nc (netcdf) and now a single file contains all the rasters for one year (365 days), and sometimes for multiple variables also.

@@

### Norwegian standard

@@

The reference condition for climate variables is defined as the last normal period 1961-1990.

@@

-   Un-scaled indicator value = median value over 5 year periods

@@

-   Thresholds for good ecosystem condition = 2 standard deviation units away from the upper reference level for the climate variable during the reference period.

-   Lower reference values (two-directional) = 5 standard deviation units for the climate variable during the reference period (implies linear scaling).

@@

For the indicator map (1 x 1 km raster) there is no uncertainty associated with the indicator values. For aggregated indicator values (e.g. for regions), the uncertainty in the indicator value is calculated from the spatial variation in the indicator values via bootstrapping. This might, however, be changed later to the temporal variation between the five years of each period.

@@

*rr and tm are being download from:* <https://thredds.met.no/thredds/catalog/senorge/seNorge_2018/Archive/catalog.html>

@@

```{r, fig.cap="Table explaining the available climate parameters in the data set."}
senorgelayers <- read_delim("data/senorgelayers.txt", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
DT::datatable(senorgelayers)
```

@@

I know the CRS, so setting it manually. Although I cannot rule out 32633, I don't think it matters.

@@

@@

Initially the data had four attributes, but I subsettet to only include *tg*.

@@

Saving these to file.

@@

Here is some dummy data for accounting areas (regions) and ecosystem occurrences as well, in case I need them later.

@@

```{r, message=FALSE}
nor <- sf::st_read("data/outlineOfNorway_EPSG25833.shp")
```

Remove marine areas from regions

@@

```{r, fig.cap="Five accounting areas (regions) in Norway."}

@@

The climate indicators need to be masked with ecosystem type maps. This step is part of this chapter.

@@

### Conceptual workflow

The general, the conceptual workflow is like this:

@@

    -   Import .nc files (loop though year 1961-1990) and subset to the correct attribute

    -   Filter data by dates (optional) (`dplyr::filter`). *This means reading all 365 rasters into memory, and it is much quicker to filter out the correct rasters in the importing step above (see examples later in this chapter)*

    -   Aggregate across time within a year (`stars::st_apply`). *This is the most time consuming operation in the workflow.*

@@

4.  Mask by ecosystem type (*This could also be done in step one, but it doesn't speed things up to set some cells to NA*)

5.  Aggregate in space (to accounting areas) (*zonal statistics*)

@@

    -   Intersect with accounting area polygons `exactextrar::exact_extract` and get mean/median and (spatial) sd. (*Alternatively, get temporal sd at the grid cell level in the step above.*)

6.  Make trend figure and spatially aggregated maps,

@@

*Later I discover that \`dplyr::filter\` work after reading the whole object to file, and therefor it is too slow for our use. I therefore filter data a different way further down.*

@@

For this example I want to calculate the mean temperature over the summer. I therefore need to aggregate over time. Many climate indicators will need this functionality. There are two ways, either using `st_apply` or using `aggregate`.

@@

The *v\_* follows the naming convention we have developed for ourselves.

@@

```{r, fig.cap="Temporal aggregation on the dummy data set."}
ggplot()+
  geom_stars(data = summerTemp_fullSeries) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~Year)
```

@@

*I initially tried to read stars proxy object and add steps to the call_list and execute these with `st_as_stars`, but this still returned proxy objects. Therefore I read the whole file in without proxy. To save time I can subset based on Julian dates.*

I will use [parallelization](https://tmieno2.github.io/R-as-GIS-for-Economists/EE.html) to speed things up.

@@

myFiles <- list.files(path2, pattern=".nc$",full.names = T)
# The last file (the last year) is incomplete and don't include all julian dates that we select below, so I will not include it:
myFiles <- myFiles[-length(myFiles)]

real_temp_summer <- NULL

# set up parallel cluster using 10 cores
cl <- makeCluster(10L)

# Get julian days after defining months
temp <- stars::read_ncdf(paste(myFiles[1]), var="tg")
start_month_num <-  6
end_month_num <- 8

@@

  tic("init")
  temp <- stars::read_ncdf(paste(myFiles[i]), var="tg", proxy=F,
                           ncsub = cbind(start = c(1, 1, julian_start), 
                              count = c(NA, NA, step)))
  year_temp <- year(st_get_dimension_values(temp, "time")[1])
  print(year_temp)
  lookup <- setNames("mean", paste0("v_", year_temp)) 
    # Perhaps leave out the v_ to get a numeric vector instead, 
    # which is easier to subset
  st_crs(temp) <- 25833
  toc()

@@

This takes about 20 sec per file/year, or 22 min on total. That is not too bad. About 6000 raster are read into memory. Here's a test for the effect of splitting over more cores.

```{r, eval=F}
cl <- makeCluster(2L)
cl2 <- makeCluster(6L)
cl3 <- makeCluster(10L)

@@

-   No cluster: 21.401 sec elapsed

-   Two clusters: 21.546 sec elapsed

-   Six clusters: 14.548 sec elapsed

-   Ten clusters: 15.29 sec elapsed

-   Ten clusters (second run): 13.617 sec elapsed

@@

These files are about 500 MB, but the tiff file reads in a lot quicker then the rds file, so from now on I'll just export as tiff. *(This could perhaps have been solved with a `gc()`,. Later I save as RData with no problems).*

@@

Note that GTiff automatically renames the third dimension *band* and also renames the attribute.

@@

I can rename them.

```{r}
temp_median_summer <- temp_median_summer %>% 
  st_set_dimensions(names = c("x", "y", "v_YEAR")) %>%
  setNames("temperature")
dim(temp_median_summer)
```

@@

### 2. Calculate reference values

#### Aggregate across reference years

@@

```{r}
st_get_dimension_values(y, "time")
```

```{r}
y_filtered <- y %>%
  select(Attribute_A) %>%
  filter(time %within% interval("1961-01-01", "1990-12-31"))
```

@@

This is also something I can export.

```{r, eval=F}
# write_stars can only export a single attribute 
#write_stars(temp_median_summer_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.tiff")

@@

system.time(
y_var_norm <- y_var %>%
  mutate(reference_low = reference_upper - 5*sd ) %>%
  mutate(reference_low2 = reference_upper + 5*sd ) %>%
  mutate(threshold_low = reference_upper -2*sd ) %>%
  mutate(threshold_high = reference_upper +2*sd ) %>%
  mutate(across(all_of(cols), ~ 
                  if_else(.x < reference_upper,
                  if_else(.x < threshold_low, 
                                        (.x - reference_low) / (threshold_low - reference_low),
                                        (.x - threshold_low) / (reference_upper - threshold_low),
                                        ),
                  if_else(.x > threshold_high, 
                                        (reference_low2 - .x) / (reference_low2 - threshold_high),
                                        (threshold_high - .x) / (threshold_high - reference_upper),
                                        )
                ))) %>%
  mutate(across(all_of(cols), ~ if_else(.x > 1, 1, .x))) %>%
  mutate(across(all_of(cols), ~ if_else(.x < 0, 0, .x))) %>%
  rename(all_of(cols_new)) %>%
  c(select(y_var, all_of(cols)))
)
```

user system elapsed 14.803 2.717 17.512

@@

The *real* indicator values should be running means over 5 years, but calculating a running mean for all time steps is too time consuming. Therefore, the time series will be presented with yearly resolution, and the scaled, regional indicator values will be calculated for distinct time steps.
